The Importance of Initial Conditions in Seasonal Predictions of Antarctic Sea Ice

### 2025-11-19 analysis/paper/access-ice.tex
Render

\raggedbottom
\usepackage[section]{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Forecast Model},
  pdfauthor={Elio Campitelli; Ariaan Purich; Julie Arblaster; Eun-Pa Lim; Matthew C. Wheeler; Phillip Reid},
  pdfkeywords={sea ice, seasonal predictability, initial
conditions, forecasting},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Evaluating the Importance of Initial Conditions for Antarctic Sea
Ice Seasonal Predictability with a Fully Coupled Forecast
Model]{Evaluating the Importance of Initial Conditions for Antarctic Sea
Ice Seasonal Predictability with a Fully Coupled Forecast Model}

% author setup
\author*[1,2]{\fnm{Elio} \sur{Campitelli}}\email{elio.campitelli@monash.edu}\author[1,2]{\fnm{Ariaan} \sur{Purich}}\author[1,2]{\fnm{Julie} \sur{Arblaster}}\author[3]{\fnm{Eun-Pa} \sur{Lim}}\author[3]{\fnm{Matthew C.} \sur{Wheeler}}\author[3]{\fnm{Phillip} \sur{Reid}}
% affil setup
\affil[1]{\orgdiv{School of Earth, Atmosphere and
Environment}, \orgname{Monash University, Australia,}}
\affil[2]{\orgdiv{Securing Antarctica's Environmental
Future}, \orgname{Monash University, Australia}}
\affil[3]{\orgdiv{Research, Bureau of Meteorology, Australia}}

@@

\abstract{Accurate Antarctic sea-ice forecasts are crucial for climate
monitoring and operational planning, yet they remain challenging due to
model biases and complex ice-ocean-atmosphere interactions. The two
versions of the Australian Bureau of Meteorology's ACCESS seasonal
forecast system, ACCESS-S1 and ACCESS-S2, use identical model
configuration and differ only in their initial conditions; primarily in
that ACCESS-S2 does not assimilate sea-ice observations, whereas
ACCESS-S1 does.\\
This provides an convenient opportunistic experiment to assess the role
of initial conditions on Antarctic sea-ice forecasts using more than 20
years of fully coupled simulations with two 9-member ensembles. Our
analysis reveals that both systems experience an extended melt season
and delayed growth phase compared with observations. This leads to a
significant negative sea-ice extent bias, which is corrected only in
ACCESS-S1 by the data assimilation system. The impact of the differing
initial conditions on forecast errors varies dramatically by season:
summer and autumn initial conditions (January-April) provide predictive
skill for up to three months, with February initial conditions being
particularly crucial. In contrast, winter forecasts of the two systems
are statistically indistinguishable after only two weeks. Regional
analysis of forecast skill suggests that this winter predictability
barrier is most dramatic over East Antarctica, where even ACCESS-S1
shows negative skill. These findings highlight the critical importance
of comprehensive year-round sampling in predictability studies and
suggest that operational sea-ice data assimilation efforts should
prioritize the summer-autumn period when initial conditions have maximum
impact on forecast skill.}

@@

ACCESS-S2 \citep{wedd2022} is the Bureau of Meteorology's seasonal
forecast system which became operational in October 2021, replacing the
ACCESS-S1 system \citep{hudson2017}. The model components of both
ACCESS-S2 and ACCESS-S1 are identical with the same numbers of levels
and resolution. They consist of the Global Atmosphere 6.0 (GA6)
\citep{williams2015, waters2017}, the Unified Model's Global Land 6.0
\citep{best2011, waters2017}, NEMO Global Ocean 5.0
\citep{gurvan2013, megann2014} and Global Sea Ice 6.0 {[}CICE;
\citet{rae2015}{]}. The atmosphere has a N216 horizontal resolution
(\textasciitilde60 km in the mid-latitudes) with 85 vertical levels. The
land model uses the same horizontal grid as the atmosphere with four
soil levels. The ocean component has a nominal horizontal resolution of
1/4º with 75 vertical levels. The sea-ice component, based on CICE
version 4.1, has the same resolution as the ocean component and five
sea-ice thickness categories as well as an open water category.

Both systems take atmospheric initial conditions derived from
ERA-interim \citep{dee2011} for their hindcasts. The main difference
between the hindcasts of the two systems are the ocean and sea-ice
initial conditions. ACCESS-S1's ocean and sea-ice initial conditions
come from the Met Office FOAM system, which uses a multivariate,
incremental three-dimensional variational (3D-Var),
first-guess-at-appropriate-time (FGAT) data assimilation scheme
\citep{waters2015} and assimilates sea surface temperature (SST), sea
surface height (SSH), in situ temperature and salinity profiles, and
satellite observations of sea-ice concentration using the EUMETSAT
OSISAF product described in the next section. ACCESS-S2, on the other
hand, is initialised from ocean conditions generated by the BoM weakly
coupled ensemble data assimilation scheme described in \citet{wedd2022}.
This scheme uses an optimal interpolation method and assimilates
temperature and salinity profiles from EN4 \citep{good2013}. SSTs are
nudged to Reynolds OISSTv2.1 \citep{reynolds2007} in areas where SSTs
are over 0ºC and Sea Surface Salinity is weakly nudged to the World
Ocean Atlas 2013 climatology \citep{zweng2013}.

@@

The ACCESS-S1 hindcast set is made up of nine members created by
perturbing the atmospheric fields only with a random field perturbation
\citep{hudson2017} and runs for 217 days for the period 1990--2012
initialised at the first of every month. The ACCESS-S2 hindcast set used
in this study runs for the period 1981--2018. Each hindcast consists of
nine ensemble members built from three consecutive three-member
forecasts created in the same manner as ACCESS-S1 members initialised at
the first of every month and the two previous days and run for 279 days.
We analyse the ensemble mean hindcasts unless otherwise specified.

@@

For verification we use satellite-derived sea-ice concentration, which
estimates the proportion of each grid area that is covered with ice.
Datasets derived using different algorithms and satellite platforms,
each have their own biases and uncertainties. Estimates of inter-product
uncertainty of sea-ice extent (SIE, defined here as the total region of
the Southern Ocean with at least 15\% sea-ice cover) is of the order of
0.5 million \(km^2\) \citep{meier2019}. As will be shown below, this
spread is minimal compared with the typical errors in the ACCESS-S2 and
ACCESS-S1 forecasts, so the overall conclusions of this study are
independent of the verification dataset used.

We use NOAA/NSIDC's Climate Data Record V4 {[}CDR; \citet{meier2014}{]}
as the primary sea-ice verification dataset. It takes the maximum value
of the NASA Team \citep{cavalieri1984} and NASA Bootstrap
\citep{comiso2023} sea-ice concentration products to reduce their low
concentration bias \citep{meier2014, meier2021}. Both source algorithms
use data from the Scanning Multichannel Microwave Radiometer (SMMR) on
the Nimbus-7 satellite and from the Special Sensor Microwave/Imager
(SSM/I) sensors on the Defense Meteorological Satellite Program's (DMSP)
-F8, -F11, and -F13 satellites. The data has a spatial resolution of 25
by 25 km and daily from November 1978 onwards.

@@

Forecasts errors are also compared with hypothetical forecasts based on
the persistence of anomalies and on climatology. The persistence
forecast is generated by extending the observed anomalies in time.

@@

Where \(RMSE_{f}\) is the RMSE of the forecast, \(RMSE_{r}\) is the RMSE
of the reference forecast. Negative skill score indicates that the
forecast is worse than the reference forecast while positive values
indicate an improvement. A perfect forecast, would have zero RMSE and
thus a skill score of 1.

@@

We performed all analysis in this paper using the R programming language
\citep{rcoreteam2020}, using data.table \citep{dowle2020} and metR
\citep{campitelli2020b} packages. Significant processing was performed
using the CDO command line operators \citep{schulzweida2023}. All
graphics are made using ggplot2 \citep{wickham2009}. The paper was
rendered using knitr and quarto \citep{xie2015, allaire2022}.

@@

\caption{\label{fig-hindcast-extent}Row a: Pan-Antarctic mean sea-ice
extent for all hindcasts initialised on the first of each calendar month
for ACCESS-S1 (column 1; green) and ACCESS-S1 (column 2; purple).
Observed mean sea-ice extent in each corresponding hindcast period is
shown in black. Row b: Mean differences between the forecast and the
observed values. Circles represent the initial conditions at the start
of forecasts (i.e., the first of every month), and triangles represent
the mean values at the end of forecasts (i.e., the longest possible lead
time).}

@@

Figure~\ref{fig-hindcast-extent} shows mean sea-ice extent of the
ACCESS-S1 and ACCESS-S2 hindcasts (row a) and their differences from
mean sea-ice extent of NOAA/NSIDC CDRV4 (row b). Mean extent at the
first of every month is indicated with circles for the initial
conditions and with triangles for the longest lead time possible for
each model (between 274 and 277 days for ACCESS-S2 and between 213 and
216 days for ACCESS-S1). At this long lead time, information of the
initial conditions is essentially lost and the forecast reverts to each
model's preferred equilibrium state.

@@

ACCESS-S1 has a comparatively smaller overall bias
(Figure~\ref{fig-bias-2}). The largest values are found between April
and June, when the faster growth results in large positive bias along
the sea-ice edge, and in January, when the faster melt leads to large
negative bias in the Weddell and Admunsen Seas.

@@

To assess ACCESS-S2 forecasts in more detail, we compute error measures
for all hindcasts started on the 1st of every month.

@@

\caption{\label{fig-rmse}Mean RMSE of sea-ice concentration anomalies as
a function of forecast lead time for all forecasts initialised on the
first of each month compared with a reference forecast of persistence of
anomalies (black) and climatology (gray). Only the first 120 days are
shown. In parenthersis, the shortest time at which ACCESS-S1 and
ACCESS-S2 mean RMSE is not statistically different at the 99\%
confidence level.}

@@

Figure~\ref{fig-rmse} shows the mean RMSE of sea-ice concentration
anomalies for ACCESS-S1 and ACCESS-S2 hindcasts compared against
persistence and climatological forecasts used as a benchmark. Due to
errors in the initial conditions, it is expected that persistence
forecasts would be better than the model forecasts at very short lead
times, but that the persistence forecast errors would grow faster and
may eventually surpass the model forecast errors. The black line shows
that the persistence forecast error indeed grows rapidly and reaches its
maximum in about 30 days for most months except for February, when it
grows much slower. The ACCESS-S1 forecast errors grow slower than
persistence forecast errors and remain lower after less than 10 days on
average. The ACCESS-S2 forecast error starts high in all months and is
lower than the persistence forecast error after more than 15 days in
most months except for forecast initialised in February, when it takes
80 days.

@@

\pandocbounded{\includegraphics[keepaspectratio]{access-ice_files/figure-pdf/fig-rmse_lon-1-1.pdf}}

@@

\caption{\label{fig-rmse_lon-1}RMSE skill score of ACCESS-S1 forecasts
with climatological forecast as reference computed on 15 meridional
slices 24º wide as a function of lead time and longitude. Antarctica's
coastline is shown at the bottom of each panel for reference.}

@@

\pandocbounded{\includegraphics[keepaspectratio]{access-ice_files/figure-pdf/fig-rmse_lon-2-1.pdf}}

@@

ACCESS-S2 forecasts (Figure~\ref{fig-rmse_lon-2}) also have lower skill
over East Antarctica. From July to December even though the
pan-Antarctic average skill is negative at all lead times
(Fig.~\ref{fig-lead-time-window}), it is positive for up to a month in
West Antarctica. Since oceanic and atmospheric forcing is the only
source of information, this suggests that sea-ice in this region is
particularly sensitive to oceanic and atmospheric forcing and suggests a
role of the Pacific-South American mode and the Amundsen Sea Low to
shape sea-ice concentration anomalies. The fact that this is evident in
the months in which El Niño--Southern Oscillation teleconnections are
more important for atmospheric circulation also suggest the influence of
tropical Pacific variability. February and March are the only two months
that can be forecasted with marginally positive skill in large regions.

@@

Ensemble spread grows quickly even when perturbations are only
implemented in the atmosphere component (in ACCESS-S1), indicating that
sea ice is indeed responding quickly to atmospheric perturbations.
However, our analysis suggests that the atmosphere and ocean data
assimilation implemented in ACCESS-S2 is only effectively influencing
sea-ice initial conditions from June to October, while the rest of the
year, the sea-ice component runs virtually free, reverting to its biased
equilibrium state.

@@

Although ACCESS-S1 only assimilates sea-ice concentration, it is clear
that sea-ice thickness is also affected through the assimilation
process. ACCESS-S1 simulates significantly thicker ice than ACCESS-S2
and in both systems sea-ice is thicker at shorter lead times than at
longer lead times. Both the explicit data assimilation in ACCESS-S2 and
the effects of atmospheric and oceanic data assimilation in ACCESS-S1
might be nudging simulated sea ice to be thicker than the model
equilimbrium state. We suggest that the thinner sea ice in ACCESS-S2
contributes to the large sea-ice extent variance, but other mechanisms,
such as unbalanced initial conditions might also be important.

Given that ACCESS-S2 sea-ice extent is not directly initialised by
sea-ice observations, comparing its forecasts with those of ACCESS-S1
allows us to estimate the time-scale over which initial conditions are
important. We find that initial conditions affect Antarctic sea-ice
forecasts in the order of a few months, but that effect is seasonally
dependent. January to April initial conditions improve forecasts for up
to three months with, February initial conditions in particular are
shown to be crucial for determining sea-ice evolution at least up to
May. Arctic sea-ice forecasts also show greater sensitivity to initial
conditions in boreal summer, compared with boreal winter
\citep{day2014, bunzel2016}, suggesting a similar mechanism might be
playing a role.

Forecasts initialised in winter have very little skill and ACCESS-S1 and
ACCESS-S2 forecast errors are not statistically different after just two
weeks. This is consistent with the findings of \citet{libera2022}'s
finding of a ``winter predictability barrier'' in the Weddell Sea,
although they describe the barrier as sharp loss of predictability in
July, and here we find a gradual reduction in skill compared with
climatology around June. This difference might be due to our use of
pan-Antarctic RMSE, since our regional analysis indicates that the
degraded skill is most dramatic in the King Haakon sea.

@@

@@

The underlying code for this study is available on github and can be
accessed via this link https://github.com/eliocamp/access-s2\_ice-eval.

Raw data of ACCESS-S1 and ACCESS-S2 forecast are not available due to
size. Derived datasets required to reproduce the results are available
in this zenodo repository: XXXXX

@@

\caption{\label{fig-hindcast-extent-osi}Row a: Pan-Antarctic mean
sea-ice extent for all hindcasts initialised on the first of each
calendar month for ACCESS-S1 (column 1; green) and ACCESS-S1 (column 2;
purple). Observed mean sea-ice extent in each corresponding hindcast
period is shown in black. Row b: Mean differences between the forecast
and the observed values. Circles represent the initial conditions at the
start of forecasts (i.e., the first of every month), and triangles
represent the mean values at the end of forecasts (i.e., the longest
possible lead time).}

@@

\caption{\label{fig-rmse-osi}Mean RMSE of sea-ice concentration
anomalies as a function of forecast lead time for all forecasts
initialised on the first of each month compared with a reference
forecast of persistence of anomalies (black) and climatology (gray).
Only the first 120 days are shown. In parenthersis, the shortest time at
which ACCESS-S1 and ACCESS-S2 mean RMSE is not statistically different
at the 99\% confidence level.}

@@

\pandocbounded{\includegraphics[keepaspectratio]{access-ice_files/figure-pdf/fig-rmse_lon-1-osi-1.pdf}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{access-ice_files/figure-pdf/fig-rmse_lon-2-osi-1.pdf}}

>>>>

\raggedbottom
\usepackage[section]{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The Importance of Initial Conditions in Seasonal Predictions of Antarctic Sea Ice},
  pdfauthor={Elio Campitelli; Ariaan Purich; Julie Arblaster; Eun-Pa Lim; Matthew C. Wheeler; Phillip Reid},
  pdfkeywords={sea ice, seasonal predictability, initial
conditions, forecasting},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[The Importance of Initial Conditions in Seasonal Predictions of
Antarctic Sea Ice]{The Importance of Initial Conditions in Seasonal
Predictions of Antarctic Sea Ice}

% author setup
\author*[1,2]{\fnm{Elio} \sur{Campitelli}}\email{elio.campitelli@monash.edu}\author[1,2]{\fnm{Ariaan} \sur{Purich}}\author[1,2]{\fnm{Julie} \sur{Arblaster}}\author[3]{\fnm{Eun-Pa} \sur{Lim}}\author[3]{\fnm{Matthew C.} \sur{Wheeler}}\author[3]{\fnm{Phillip} \sur{Reid}}
% affil setup
\affil[1]{\orgdiv{School of Earth, Atmosphere and
Environment}, \orgname{Monash University, Kulin Nations, Clayton,
Victoria, Australia.}}
\affil[2]{\orgdiv{SARC Special Research Initiative for Securing
Antarctica's Environmental Future}, \orgname{Clayton, Kulin Nations,
Victoria, Australia}}
\affil[3]{\orgdiv{Research, Bureau of Meteorology, Australia}}

@@

\abstract{Accurate Antarctic sea-ice forecasts are crucial for climate
monitoring and operational planning, yet they remain challenging due to
model biases and complex ice-ocean-atmosphere interactions. The two
versions of the Australian Bureau of Meteorology's ACCESS seasonal
forecast system, ACCESS-S1 and ACCESS-S2, use identical model
configuration and differ only in their initial conditions; primarily in
that ACCESS-S2 does not assimilate sea-ice observations, whereas
ACCESS-S1 does.\\
This provides a convenient opportunistic experiment to assess the role
of initial conditions on Antarctic sea-ice forecasts using more than 20
years of fully coupled simulations with two 9-member ensembles. Our
analysis reveals that both systems experience an extended melt season
and delayed growth phase compared with observations. This leads to a
significant negative sea-ice extent bias, which is corrected only in
ACCESS-S1 by the data assimilation system. The impact of the differing
initial conditions on forecast errors varies dramatically by season:
summer and autumn initial conditions (January-April) provide predictive
skill for up to three months, with February initial conditions being
particularly crucial. In contrast, winter forecasts of the two systems
are statistically indistinguishable after only two weeks. Regional
analysis of forecast skill suggests that this winter predictability
barrier is most dramatic over East Antarctica, where even ACCESS-S1
shows negative skill. These findings highlight the critical importance
of comprehensive year-round sampling in predictability studies and
suggest that operational sea-ice data assimilation efforts should
prioritize the summer-autumn period when initial conditions have maximum
impact on forecast skill.}

@@

ACCESS-S2 \citep{wedd2022} is the Bureau of Meteorology's seasonal
forecast system which became operational in October 2021, replacing the
ACCESS-S1 system \citep{hudson2017}. The model components of both
ACCESS-S2 and ACCESS-S1 are identical with the same numbers of levels
and resolution. They consist of the Global Atmosphere 6.0 (GA6)
\citep{williams2015, waters2017}, the Unified Model's Global Land 6.0
\citep{best2011, waters2017}, NEMO Global Ocean 5.0
\citep{gurvan2013, megann2014} and Global Sea Ice 6.0 {[}CICE;
\citet{rae2015}{]}. The atmosphere has a N216 horizontal resolution
(\textasciitilde60 km in the mid-latitudes) with 85 vertical levels. The
land model uses the same horizontal grid as the atmosphere with four
soil levels. The ocean component has a nominal horizontal resolution of
1/4° with 75 vertical levels. The sea-ice component, based on CICE
version 4.1, has the same resolution as the ocean component and five
sea-ice thickness categories as well as an open water category.

Both systems take atmospheric initial conditions derived from
ERA-interim \citep{dee2011} for their hindcasts. The main difference
between the hindcasts of the two systems are the ocean and sea-ice
initial conditions. ACCESS-S1's ocean and sea-ice initial conditions
come from the Met Office FOAM system, which uses a multivariate,
incremental three-dimensional variational (3D-Var),
first-guess-at-appropriate-time (FGAT) data assimilation scheme
\citep{waters2015} and assimilates sea surface temperature (SST), sea
surface height (SSH), in situ temperature and salinity profiles, and
satellite observations of sea-ice concentration using the EUMETSAT
OSISAF product described in the next section. ACCESS-S2, on the other
hand, is initialised from ocean conditions generated by the BoM weakly
coupled ensemble data assimilation scheme described in \citet{wedd2022}.
This scheme uses an optimal interpolation method and assimilates
temperature and salinity profiles from EN4 \citep{good2013}. SSTs are
nudged to Reynolds OISSTv2.1 \citep{reynolds2007} in areas where SSTs
are over 0°C and Sea Surface Salinity is weakly nudged to the World
Ocean Atlas 2013 climatology \citep{zweng2013}.

@@

The ACCESS-S1 hindcast set is made up of nine members created by
perturbing the atmospheric fields only with a random field perturbation
\citep{hudson2017} and runs for 217 days for the period 1990--2012
initialised at the first of every month. The ACCESS-S2 hindcast set used
in this study runs for the period 1981--2018. Ensemble members are
created in the same manner as ACCESS-S1 members, however, due to
computing cost limitations, only three members per forecast
initialisation date were run for 279 days. Bigger ensembles were
generated by aggregating several three-member ensembles initialised on
successive days \citep{wedd2022}. Here, we build a nine-member
time-lagged ensemble from three consecutive three-member forecasts
initialised at the first of every month and the two previous days and
run for 279 days. We analyse the ensemble mean hindcasts unless
otherwise specified.

@@

For verification we use satellite-derived sea-ice concentration, which
estimates the proportion of each grid area that is covered with ice.
Datasets derived using different algorithms and satellite platforms,
each have their own biases and uncertainties. Estimates of inter-product
uncertainty of sea-ice extent (SIE, defined here as the total region of
the Southern Ocean with at least 15\% sea-ice cover) are of the order of
0.5 million \(km^2\) \citep{meier2019}. As will be shown below, this
spread is minimal compared with the typical errors in the ACCESS-S2 and
ACCESS-S1 forecasts, so the overall conclusions of this study are
independent of the verification dataset used.

We use NOAA/NSIDC's Climate Data Record V4 {[}CDR; \citet{meier2014}{]}
as the primary sea-ice verification dataset. It takes the maximum value
of the NASA Team \citep{cavalieri1984} and NASA Bootstrap
\citep{comiso2023} sea-ice concentration products to reduce their low
concentration bias \citep{meier2014, meier2021}. Both source algorithms
use data from the Scanning Multichannel Microwave Radiometer (SMMR) on
the Nimbus-7 satellite and from the Special Sensor Microwave/Imager
(SSM/I) sensors on the Defense Meteorological Satellite Program's (DMSP)
-F8, -F11, and -F13 satellites. The data have a spatial resolution of 25
by 25 km and daily from November 1978 onwards.

@@

Forecast errors are also compared with hypothetical forecasts based on
the persistence of anomalies and on climatology. The persistence
forecast is generated by extending the observed sea-ice concentration
anomalies the day of the forecast initialisation and comparing it with
the actual anomalies observed. The climatological forecast error is
computed as the standard deviation of daily anomalies.

@@

Where \(RMSE_{f}\) is the RMSE of the forecast, \(RMSE_{r}\) is the RMSE
of the reference forecast. Negative skill score indicates that the
forecast is worse than the reference forecast while positive values
indicate an improvement. A perfect forecast would have zero RMSE and
thus a skill score of 1.

@@

We performed all analyses in this paper using the R programming language
\citep{rcoreteam2020}, using data.table \citep{dowle2020} and metR
\citep{campitelli2020b} packages. Significant processing was performed
using the CDO command line operators \citep{schulzweida2023}. All
graphics are made using ggplot2 \citep{wickham2009}. The paper was
rendered using knitr and Quarto \citep{xie2015, allaire2022}.

@@

\caption{\label{fig-hindcast-extent}Row a: Pan-Antarctic daily mean
sea-ice extent for all hindcasts initialised on the first of each
calendar month for ACCESS-S1 (column 1; green) and ACCESS-S2 (column 2;
purple). Observed mean sea-ice extent in each corresponding hindcast
period is shown in black. Row b: Mean differences between the forecast
and the observed values. Circles represent the initial conditions at the
start of forecasts (i.e., the first of every month), and triangles
represent the mean values at the lead time corresponding to the maximum
lead time in S1 (between 213 and 216 days, depending on the month)}

@@

Figure~\ref{fig-hindcast-extent} shows mean sea-ice extent of the
ACCESS-S1 and ACCESS-S2 hindcasts (row a) and their differences from
mean sea-ice extent of NOAA/NSIDC CDRV4 (row b). Mean extent at the
first of every month is indicated with circles for the initial
conditions and with triangles for the longest lead time possible for
each model (between 274 and 277 days for ACCESS-S2 and between 213 and
216 days for ACCESS-S1). At this long lead time, information of the
initial conditions is essentially lost and the forecast reverts close to
each model's preferred equilibrium state.

@@

ACCESS-S1 has a comparatively smaller overall bias
(Figure~\ref{fig-bias-2}). The largest values are found between April
and June, when the faster growth results in large positive bias along
the sea-ice edge, and in January, when the faster melt leads to large
negative bias in the Weddell and Amundsen Seas.

@@

@@

\caption{\label{fig-rmse}Mean RMSE of sea-ice concentration anomalies as
a function of forecast lead time for all forecasts initialised on the
first of each month compared with a reference forecast of persistence of
anomalies (black) and climatology (gray). Only the first 120 days are
shown. In parentheses, the shortest time at which ACCESS-S1 and
ACCESS-S2 mean RMSE is not statistically different at the 99\%
confidence level.}

@@

To assess ACCESS-S2 forecasts in more detail, we compute error measures
for all hindcasts started on the 1st of every month.
Figure~\ref{fig-rmse} shows the mean RMSE of sea-ice concentration
anomalies for ACCESS-S1 and ACCESS-S2 hindcasts compared against
persistence and climatological forecasts used as a benchmark. Due to
errors in the initial conditions, it is expected that persistence
forecasts would be better than the model forecasts at very short lead
times, but that the persistence forecast errors would grow faster and
may eventually surpass the model forecast errors. The black line shows
that the persistence forecast error indeed grows rapidly and reaches its
maximum in about 30 days for most months except for February, when it
grows much slower. The ACCESS-S1 forecast errors grow slower than
persistence forecast errors and remain lower after less than 10 days on
average. The ACCESS-S2 forecast error starts high in all months and is
lower than the persistence forecast error after more than 15 days in
most months except for forecast initialised in February, when it takes
80 days.

@@

\pandocbounded{\includegraphics[keepaspectratio]{access-ice_files/figure-pdf/fig-rmse_lon-1-1.png}}

@@

\caption{\label{fig-rmse_lon-1}RMSE skill score of ACCESS-S1 forecasts
with climatological forecast as reference computed on 15 meridional
slices 24° wide as a function of lead time and longitude. Antarctica's
coastline is shown at the bottom of each panel for reference.}

@@

\pandocbounded{\includegraphics[keepaspectratio]{access-ice_files/figure-pdf/fig-rmse_lon-2-1.png}}

@@

ACCESS-S2 forecasts (Figure~\ref{fig-rmse_lon-2}) also have lower skill
over East Antarctica. From July to December even though the
pan-Antarctic average skill is negative at all lead times
(Fig.~\ref{fig-lead-time-window}), it is positive for up to a month in
West Antarctica. Since oceanic and atmospheric forcing is the only
source of information, this suggests that sea-ice in this region is
particularly sensitive to oceanic and atmospheric forcing and suggests a
role of the Pacific-South American mode and the Amundsen Sea Low to
shape sea-ice concentration anomalies. The fact that this is evident in
the months in which El Niño--Southern Oscillation teleconnections are
more important for atmospheric circulation also suggests the influence
of tropical Pacific variability. February and March are the only two
months that can be forecasted with marginally positive skill in large
regions.

\pandocbounded{\includegraphics[keepaspectratio]{access-ice_files/figure-pdf/fig-rmse_lon-3-1.png}}

\caption{\label{fig-rmse_lon-3}Same as Figure~\ref{fig-rmse_lon-1} but
for the difference between ACCESS-S1 and ACCESS-S2.}

Finally, Figure~\ref{fig-rmse_lon-3} shows the difference in skill
between ACCESS-S1 and ACCESS-S2. Large differences in skill indicate
areas and months that are most affected by the data assimilation present
in ACCESS-S1. Between January and March, which are the months in which
ACCESS-S1 is the most skilful (Figure~\ref{fig-lead-time-window}), most
of the improvement compared with ACCESS-S2 is present in the Ross and
Weddell Sea. In April and May, the improvement seems more homogeneous.

@@

Ensemble spread grows quickly even when perturbations are only
implemented in the atmosphere component (in ACCESS-S1), indicating that
sea ice is indeed responding quickly to atmospheric perturbations.
However, our analysis suggests that the atmosphere and ocean data
assimilation implemented in ACCESS-S2 is only effectively influencing
sea-ice initial conditions from June to October, while the rest of the
year, the sea-ice component runs virtually free, reverting to its biased
equilibrium state. \citet{zhou2022} had previously evaluated sea-ice
forecasts in ACCESS-S2 and also highlighted the poor performance of this
forecasting system attributed to the lack of good initial conditions.

@@

Although ACCESS-S1 only assimilates sea-ice concentration, it is clear
that sea-ice thickness is also affected through the assimilation
process. ACCESS-S1 simulates significantly thicker ice than ACCESS-S2
and in both systems sea-ice is thicker at shorter lead times than at
longer lead times. Both the explicit data assimilation in ACCESS-S2 and
the effects of atmospheric and oceanic data assimilation in ACCESS-S1
might be nudging simulated sea ice to be thicker than the model
equilibrium state. We suggest that the thinner sea ice in ACCESS-S2
contributes to the large sea-ice extent variance, but other mechanisms,
such as unbalanced initial conditions might also be important.

Given that ACCESS-S2 sea-ice extent is not directly initialised by
sea-ice observations, comparing its forecasts with those of ACCESS-S1
allows us to estimate the time-scale over which initial conditions are
important. We find that initial conditions affect Antarctic sea-ice
forecasts in the order of a few months, but that effect is seasonally
dependent. January to April initial conditions improve forecasts for up
to three months. February initial conditions in particular are shown to
be crucial for determining sea-ice evolution at least up to May. Arctic
sea-ice forecasts also show greater sensitivity to initial conditions in
boreal summer, compared with boreal winter \citep{day2014, bunzel2016},
suggesting a similar mechanism might be playing a role.

Forecasts initialised in winter have very little skill and ACCESS-S1 and
ACCESS-S2 forecast errors are statistically indistinguishable after just
two weeks. This is consistent with \citet{libera2022}'s finding of a
``winter predictability barrier'' in the Weddell Sea, although they
describe the barrier as a sharp loss of predictability in July, and here
we find a gradual reduction in skill compared with climatology around
June. This difference might be due to our use of pan-Antarctic RMSE,
since our regional analysis indicates that the degraded skill is most
dramatic in the King Haakon Sea.

@@

We thank the internal reviewers Bethan White and Xiaobing Zhou for their
comments and feedback.

This work benefited from earlier unpublished work by Laura Davies, Phil
Reid, Andrew G. Marshall.

This research was undertaken with the assistance of resources from the
National Computational Infrastructure (NCI Australia), an NCRIS enabled
capability supported by the Australian Government.

@@

The underlying code for this study is available on GitHub and can be
accessed via this link https://github.com/eliocamp/access-s2\_ice-eval.

Raw data of ACCESS-S1 and ACCESS-S2 forecasts are not available due to
size. Derived datasets required to reproduce the results are available
in this Zenodo repository: https://zenodo.org/records/17479538
\citep{campitelli2025}

@@

\caption{\label{fig-hindcast-extent-osi}Row a: Pan-Antarctic daily mean
sea-ice extent for all hindcasts initialised on the first of each
calendar month for ACCESS-S1 (column 1; green) and ACCESS-S2 (column 2;
purple). Observed mean sea-ice extent in each corresponding hindcast
period is shown in black. Row b: Mean differences between the forecast
and the observed values. Circles represent the initial conditions at the
start of forecasts (i.e., the first of every month), and triangles
represent the mean values at the lead time corresponding to the maximum
lead time in S1 (between 213 and 216 days, depending on the month)}

@@

\caption{\label{fig-rmse-osi}Mean RMSE of sea-ice concentration
anomalies as a function of forecast lead time for all forecasts
initialised on the first of each month compared with a reference
forecast of persistence of anomalies (black) and climatology (gray).
Only the first 120 days are shown. In parentheses, the shortest time at
which ACCESS-S1 and ACCESS-S2 mean RMSE is not statistically different
at the 99\% confidence level.}

@@

\pandocbounded{\includegraphics[keepaspectratio]{access-ice_files/figure-pdf/fig-rmse_lon-1-osi-1.png}}

@@

\pandocbounded{\includegraphics[keepaspectratio]{access-ice_files/figure-pdf/fig-rmse_lon-2-osi-1.png}}
